---
title: "Student Personas"
output:
  html_document:
    df_print: paged
---

## Section 1: Data Exploration

``` {R}
library(readr)
library(dplyr)
library(ggplot2)
library(caret)
library(cluster)
library(factoextra)
data <- read.csv("personas.csv")
summary(data)
str(data)
```

### Feature description
**Program** - Student’s declared degree pathway - Categorical 

**College** - The college division the student’s degree belongs to - Categorical

**Age** - The students age - Numerical

**Gender** - The students gender - Categorical

**FirstGen** - Indicates if the student is a first-generation college student in their family - Categorical

**Race** - The student’s ethnicity ('Hispanic/Latino' or other specified race) - Categorical

**TSI_MATH** - The student’s Texas Success Initiative Assessment score for Math - Numerical

**TSI_WRITING** - The student’s Texas Success Initiative Assessment score for Writing - Numerical

**TSI_READING** - The student’s Texas Success Initiative Assessment score for Reading - Numerical

**MotherEducation** - The highest level of education the student’s mother has attained - Categorical

**FatherEducation** - The highest level of education the student’s father has attained - Categorical

**Homeschooled** - Indicates if the student was homeschooled - Categorical **(0 - No, 1 - Yes)**

**HasGED** - Indicates if the student has their GED - Categorical **(0 - No, 1 - Yes)**

**Highschool** - The high school the student attended - Categorical

**FinAid** - The type of financial aid the student qualified for ('Other' or 'Fed Pell Grant') - Categorical

**EFC** - The Expected Family Contribution financially - Numerical

### Data exploration
``` {R}
# Age
ggplot(data, aes(x = Age)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(title = "Histogram of Age", x = "Age", y = "Count")

# Homeschooled
ggplot(data, aes(x = factor(Homeschooled))) +
  geom_bar(fill = "green") +
  labs(title = "Bar Plot of Homeschooled", x = "Homeschooled", y = "Count")

# HasGED
ggplot(data, aes(x = factor(HasGED))) +
  geom_bar(fill = "green") +
  labs(title = "Bar Plot of HasGED", x = "HasGED", y = "Count")

# Program
ggplot(data, aes(x = Program)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of Program", x = "Program", y = "Count")

# College
ggplot(data, aes(x = College)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of College", x = "College", y = "Count")

# Gender
ggplot(data, aes(x = Gender)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of Gender", x = "Gender", y = "Count")

# FirstGen
ggplot(data, aes(x = FirstGen)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of FirstGen", x = "FirstGen", y = "Count")

# Race
ggplot(data, aes(x = Race)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of Race", x = "Race", y = "Count")

# TSI_MATH
ggplot(data, aes(x = TSI_MATH)) +
  geom_density(fill = "orange") +
  labs(title = "Density Plot of TSI_MATH", x = "TSI_MATH", y = "Density")

# TSI_WRITING
ggplot(data, aes(x = TSI_WRITING)) +
  geom_density(fill = "orange") +
  labs(title = "Density Plot of TSI_WRITING", x = "TSI_WRITING", y = "Density")

# TSI_READING
ggplot(data, aes(x = TSI_READING)) +
  geom_density(fill = "orange") +
  labs(title = "Density Plot of TSI_READING", x = "TSI_READING", y = "Density")

# MotherEducation
ggplot(data, aes(x = MotherEducation)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of MotherEducation", x = "MotherEducation", y = "Count")

# FatherEducation
ggplot(data, aes(x = FatherEducation)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of FatherEducation", x = "FatherEducation", y = "Count")

# Highschool
ggplot(data, aes(x = Highschool)) +
  geom_density(fill = "orange") +
  labs(title = "Density Plot of Highschool", x = "Highschool", y = "Density")

# FinAid
ggplot(data, aes(x = FinAid)) +
  geom_bar(fill = "red") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Bar Plot of FinAid", x = "FinAid", y = "Count")

# EFC
ggplot(data, aes(x = EFC)) +
  geom_density(fill = "orange") +
  labs(title = "Density Plot of EFC", x = "EFC", y = "Density")

```

The following variables are High Cardinality Variables since they contain a large amount of distinct values. Top n frequent values helps to better visualize these.

``` {R}
# Function to display top n frequent values
top_n_frequent <- function(data, variable, n = 10) {
  freq <- as.data.frame(table(data[[variable]]))
  names(freq) <- c("Value", "Frequency")
  freq <- freq[order(-freq$Frequency), ]
  head(freq, n)
}
# TSI_MATH
top_n_frequent(data, "TSI_MATH")
# TSI_WRITING
top_n_frequent(data, "TSI_WRITING")
# TSI_READING
top_n_frequent(data, "TSI_READING")
# Highschool
top_n_frequent(data, "Highschool")
# EFC
top_n_frequent(data, "EFC")
```
### Observations
- There is a very small amount of students that have been Homeschooled. HasGED is very similar, although there are more 'Yes' values.
- There are missing values, labeled as "Unknown" for some of the features. The variables that have a high "Unknown" value are FirstGen, MotherEducation, FatherEducation.
- The following variables have a high amount of NULL values - TSI_MATH, TSI_WRITING, TSI_READING

### Data cleaning and preparation

We will leave "Unkown" to be a separate category for variables such as FirstGen, MotherEducation, FatherEducation, Gender since they represent a large portion of the dataset. It could be renamed to "Other" to better reflect the third category.

We can assume that NULLS for TSI exams represent students that have not taken the exam. We can replace NULL with 0.

``` {R}
# Replace "NULL" with 0 for TSI variables
data$TSI_MATH[data$TSI_MATH == "NULL"] <- 0
data$TSI_WRITING[data$TSI_WRITING == "NULL"] <- 0
data$TSI_READING[data$TSI_READING == "NULL"] <- 0
```

``` {R}
# TSI_MATH
top_n_frequent(data, "TSI_MATH")
# TSI_WRITING
top_n_frequent(data, "TSI_WRITING")
# TSI_READING
top_n_frequent(data, "TSI_READING")
```
The TSI variables contain scores from two versions of the TSI exams. The original TSI exam scores were less than 400. The new TSI exam scores are greater than 900. We will replace the values in those variables with 0 for "Developmental Courses Needed" or 1 for "Ready For College Courses"

``` {R}
# Function to convert values based on the specified condition
convert_to_binary <- function(x) {
  as.integer(x < 350 | (x > 400 & x < 950))
}

# Apply the function to the specified columns
data$TSI_MATH <- convert_to_binary(as.numeric(data$TSI_MATH))
data$TSI_WRITING <- convert_to_binary(as.numeric(data$TSI_WRITING))
data$TSI_READING <- convert_to_binary(as.numeric(data$TSI_READING))
```

``` {R}
# TSI_MATH
top_n_frequent(data, "TSI_MATH")
# TSI_WRITING
top_n_frequent(data, "TSI_WRITING")
# TSI_READING
top_n_frequent(data, "TSI_READING")
```

``` {R}
library(dplyr)
library(tidyr)

# Converting ordinal variables to numeric
data <- data %>% 
  mutate(
    MotherEducation = case_when(
      MotherEducation == "Associate Degree" ~ 1,
      MotherEducation == "Bachelor/Four-year Degree" ~ 2,
      MotherEducation == "Graduate/Professional Degree" ~ 3,
      MotherEducation == "High School Diploma or GED" ~ 4,
      MotherEducation == "No High School" ~ 5,
      MotherEducation == "Some College" ~ 6,
      MotherEducation == "Some High School" ~ 7,
      MotherEducation == "Unknown" ~ 8
    ),
    FatherEducation = case_when(
      FatherEducation == "Associate Degree" ~ 1,
      FatherEducation == "Bachelor/Four-year Degree" ~ 2,
      FatherEducation == "Graduate/Professional Degree" ~ 3,
      FatherEducation == "High School Diploma or GED" ~ 4,
      FatherEducation == "No High School" ~ 5,
      FatherEducation == "Some College" ~ 6,
      FatherEducation == "Some High School" ~ 7,
      FatherEducation == "Unknown" ~ 8
    ),
    FirstGen = case_when(
      FirstGen == "Unknown" ~ 3,
      FirstGen == "Yes" ~ 1,
      FirstGen == "No" ~ 0
    )
  )

# Handling binary variables
data$Homeschooled <- as.numeric(data$Homeschooled)
data$HasGED <- as.numeric(data$HasGED)

# Since there are many different high schools, we take into account the top 10 and then use a separate "Other" category for the other ones
# Calculate the top 10 most common high schools
top_schools <- data %>%
  count(Highschool, sort = TRUE) %>%
  top_n(10, n) %>%
  pull(Highschool)

# Replace less common high schools with 'Other'
data <- data %>%
  mutate(Highschool = ifelse(Highschool %in% top_schools, Highschool, "Other"))
```

``` {R}
str(data)
```
``` {R}
library(data.table)
library(mltools)

# One hot encoding variables
data_dt <- as.data.table(data)
cols_to_encode <- c("Program", "College", "Gender", "Race", "Highschool", "FinAid")
data_dt[, (cols_to_encode) := lapply(.SD, as.factor), .SDcols = cols_to_encode]
data_encoded <- one_hot(data_dt, dropUnusedLevels = TRUE)
data_encoded_df <- as.data.frame(data_encoded)

# Create a list to hold mapping for each categorical variable
mapping_list <- lapply(cols_to_encode, function(col) {
  # Get the unique levels for the categorical variable
  levels <- unique(data_dt[[col]])
  
  # Ensure consistent column names across all data frames
  mapping_df <- data.frame(
    original_value = levels,
    encoded_column = paste0(col, "_", as.character(levels))
  )
  
  return(mapping_df)
})

# Combine all the mappings into a single data frame
mapping_table <- do.call(rbind, mapping_list)
```

``` {R}
# Find the median of the EFC values excluding the non-numeric entries
good_numeric_efc <- as.numeric(data$EFC[grepl("^[0-9]+$", data$EFC)])
median_efc <- median(good_numeric_efc, na.rm = TRUE)

# Replace non-numeric values with NA (to be replaced with the median later)
data$EFC <- as.numeric(data$EFC)
data$EFC[!grepl("^[0-9]+$", data$EFC)] <- NA

# Replace NA values with the median
data$EFC[is.na(data$EFC)] <- median_efc

# Apply the same conversion to the encoded data frame
data_encoded_df$EFC <- as.numeric(data_encoded_df$EFC)
data_encoded_df$EFC[is.na(data_encoded_df$EFC)] <- median_efc

```

``` {R}
str(data_encoded_df)
```

To minimize dimensionality, apply Principal Component Analysis (PCA)

```{r}

# Standardize the data (giving it a mean of 0 and a standard deviation of 1) by applying scale to the encoded data
data_scaled <- scale(data_encoded_df)

# Perform PCA 
pca_result <- prcomp(data_scaled, center = TRUE, scale. = TRUE)

# Examine the summary to determine the number of principal components
summary(pca_result)
plot(pca_result, type = "lines")

```

Given the scree plot and the summary results:  

- The first component explains approximately 4% of the variance, and each subsequent component explains slightly less.  
- The cumulative proportion of variance explained by the first 10 components is around 26%.  
- The 'elbow' in the scree plot appears around the 4th or 5th component, indicating that additional components provide diminishing returns on variance explained.  

Based on these observations, let's start with the first 5 components for clustering.  

```{r}
# PCA based dataset with the first five components
pca_data <- pca_result$x[, 1:5]

```

Apply K-Means clustering model on PCA reduced data

```{r}

# Determine the number of clusters using the Elbow Method
fviz_nbclust(pca_data, kmeans, method = "wss") + geom_vline(xintercept = 4, linetype = 2)

```

Based on the chart, we'll start with four clusters.

```{r}
# Convert pca_data to a dataframe
pca_data <- as.data.frame(pca_result$x[, 1:5])

# Properly name the columns to reflect that they are principal components
colnames(pca_data) <- c("PC1", "PC2", "PC3", "PC4", "PC5")

# Perform K-means clustering
set.seed(123)
kmeans_result <- kmeans(pca_data, centers = 4)

# Add the cluster assignments to the PCA data
pca_data$cluster <- kmeans_result$cluster

# Visualize the clusters with a scatter plot for the first two principal components
library(ggplot2)
ggplot(pca_data, aes(PC1, PC2, color = as.factor(cluster))) +
  geom_point(alpha = 0.5) +
  labs(color = 'Cluster') +
  theme_minimal()

# Visualize clustering in 3D using the first three principal components
library(rgl)
plot3d(pca_data$PC1, pca_data$PC2, pca_data$PC3, col = kmeans_result$cluster)

```

We now need to interpret the results.

```{r}
# Define the Mode function
Mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

# Add cluster assignments to the original data
data$cluster <- kmeans_result$cluster

# Decoding the one-hot encoded variables
decoded_clusters <- lapply(1:max(data$cluster), function(cluster_number) {
  cluster_data <- data[data$cluster == cluster_number, ]
  sapply(cols_to_encode, function(col) {
    # Filter original values that are present in the cluster data
    # original_values <- cluster_data[col]
    # Return the mode of the original values
    Mode(cluster_data[[col]])
  })
})

# Convert the list to a dataframe for easier viewing
decoded_cluster_modes <- do.call(rbind, decoded_clusters)

# Assign column names to match the original categorical variables
colnames(decoded_cluster_modes) <- cols_to_encode

# `decoded_cluster_modes` will give the most frequent original values of the categorical variables for each cluster
print(decoded_cluster_modes)

```

***Cluster 1:*** This cluster is characterized mostly by individuals in the Business Administration program from the Business college. The majority are female and identify as Hispanic/Latino. The most common high school is MISD- LHS, and the financial aid category is 'OTHER'.

***Cluster 2:*** Members of this cluster are predominantly in the Engineering program, associated with the Career and Technical college. They are mostly male and also identify as Hispanic/Latino. Like Cluster 1, the most common high school is MISD- LHS, and the financial aid category is 'OTHER'.

***Cluster 3:*** This cluster primarily consists of individuals in the Nursing-Vocational program from the Health Science college. They are mostly female, identify as Hispanic/Latino, but unlike the first two clusters, the most common high school category is 'Other', indicating a more diverse background in terms of high schools. The financial aid category is also 'OTHER'.

***Cluster 4:*** Similar to Cluster 3, this cluster is characterized by individuals predominantly from the Career and Technical college and in the Education program. The majority are female and identify as Hispanic/Latino, with a high school category of 'Other', and financial aid category as 'OTHER'.